Reproducibility of SNP calling in multiple sequencing runs from single tumors

### 2014-10-28 PAPER/paper_d1.tex
more paper drafts

A) What is the general field of study?
Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor, with a median survival of 13.9 months and a 5-year survival of ~5\% \cite{TCGA-GBM-13}. Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemoteraputics. Like all cancers, GBM is an evolutionary disease caused by mutations and other alterations to normal glial cell lines. However, following substantial research efforts dedicated to identifying all of the mutations in an individual patient's tumor to try and identify causal variants, few precise causal elationships have been identified. There are likely many related causes for this failure, including but not limited to genomic heterogeneity in tumors, \cite{ValentineD}, the view that "GBM growth is driven by a signaling network with functional redundancy that permits adaptation" \cite{TCGA-GBM-13}, and imperfect computational methods. 
%B) What is the specific question of the research?
In this paper, we are interested in the third. We used technical sequencing replicates from 55 GBM tumors to look at the reproducbility of the raw data using a standard computational pipeline. Assuming that those mutations that appear in each of the technical replicates are true positives, and all others are false positives (which in fact may or may not be true), we want to see what proportion of mutations recovered are ture positives, and whether or not the proportion is increased by SNP-calling methods designed to remove false positives.

@@

%B) Put your results into perspective, discussing (i) what they imply and (ii) how they compare to existing literature
GBM, primary brain cancer, is an evolutionary process by which mutaions arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism of the whole. Treatment for GBM is difficult because complete surgical resection is impossible, and recurrence is a near certainty. Even the best treatment, surgical resection followed by readiation and chemoterapy, extends life expectency by less than a year and results in decreases in quality of life for pateints \cite{GBM-treatments}. Creation of better therapies is essential, but will require a more complete understanding of the mutational and physiological landscape of GBM.

%C) Address potential shortcomings. What could have gone wrong, which parts of the research might be misleading, are there any other caveats?
-- using repatability is also problematic 1) cancer is heterogenous; just becuase a mutaiton doesn't show up in the replicate doesn't mean it's not real
-- using repatability is also problematic 2) could be degraded DNA, so large number of mutations could still be false positive
-- we may have shown that filtering doesn't do very much good, but we haven't really given alternatives
-- LOH and VAQ each remove a binch of probably-real signal, but they remove a bunch of false signal too
--

%D) (optional) Describe implications for future research; are there obvious extensions?
-- other snp callers
-- biological replicates
-- our own data?

@@

\begin{figure}
\includegraphics[scale=1.0]{C282_v_C484.png}
\caption{This figure plots the number of putative SNPs called by SomaticSniper (unfiltered) using the WGS v. WGA. Each point is a patient. The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
\end{figure}

\begin{figure}
\includegraphics[scale=1.0]{unfiltered_overlap_WGS_WGA_together_densities.png}
\caption{This figure shows the density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other sample from the same patient. The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter. }
\end{figure}

\begin{figure}
\includegraphics[scale=1.0]{unfiltered_total_muts_v_percent_overlap.png}
\caption{Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample}
\end{figure}

\begin{figure}
\includegraphics[scale=1.0]{boxplot_number_filtered.png}
\caption{Boxplot of the total number of putative SNPs removed by each of six filters, in each of 379 sequencing samples, from 311 patients. The x-axis gives the name of those filters that removed putative SNPs, and the y-axis gives, on a log scale, the number of mutations removed by a given sample.}
\end{figure}

\begin{figure}
\includegraphics[scale=1.0]{boxplot_percent_overlap_filtered.png}
\caption{Boxplot of the percentage of putative SNPs making up the overlap between the WGS and WGA sequences of a single sample that was removed by a given filter. The x-axis describes the filters, and the y-axis gives the percentage of the overlap removed.}
\end{figure}

>>>>

%A) What is the general field of study?
Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor, with a median survival of 13.9 months and a 5-year survival of ~5\% \cite{TCGA-GBM-13}. Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemoteraputics. Like all cancers, GBM is an evolutionary disease caused by mutations and other alterations to normal glial cell lines. However, following substantial research efforts dedicated to identifying all of the mutations in an individual patient's tumor to try and identify causal variants, few precise causal elationships have been identified. There are likely many related causes for this failure, including but not limited to genomic heterogeneity in tumors, \cite{ValentineD}, the view that "GBM growth is driven by a signaling network with functional redundancy that permits adaptation" \cite{TCGA-GBM-13}, and imperfect computational methods. 
%B) What is the specific question of the research?
In this paper, we are interested in the third. We used technical sequencing replicates from 55 GBM tumors to look at the reproducbility of the raw data using a standard computational pipeline. Assuming that those mutations that appear in each of the technical replicates are true positives, and all others are false positives (which in fact may or may not be true), we want to see what proportion of mutations recovered are ture positives, and whether or not the proportion is increased by SNP-calling methods designed to remove false positives.

@@

%B) Put your results into perspective, discussing (i) what they imply and (ii) how they compare to existing literature
GBM, primary brain cancer, is an evolutionary process by which mutaions arise in glial cell lines and these mutated cells and their lineages co-opt the surrounding tissue and systems to the detriment of the organism of the whole. Treatment for GBM is difficult and has poor outcomes \cite{GBM-treatments}, but could hopefully be improved by a more complete understanding of the disease. Large-scale genomics platforms and big data may be one route towards improved understanding of this devestating condition, but there are significant limitations. First (does the data look like the tumor)? Second, (PCR is long and expensive to do for every single potential point mutation). (We make a first pass at evalutating these limitations, with an eye towards maybe fixing them in the future).   
%Discuss 2-4 concrete elements of the data:
% - high numer of mutations in unfiltered case (which seems to be real, or at least repeatable)
The high number of mutations in some, but not all, of the TCGA GBM samples is, at a minimum, repeatable across technical replicates. Further, in samples with higher frequency of mutations, technical replicates have as great a percent overlap of individual putative mutations as found in less frequently-mutated samples. At a minimum, this suggests the possibility that a higher mutation rate could be a feature of a subset of GBM tumors, and not only a data artifact.

%- effects of various filters
%- discuss figure of what was removed from one filter vs. the other
%-- maybe better off with raw data? minus the filters that work
When comparing the effects of the various standard filters we employed to examine the data, only five filters of eight removed an appreciable number of putative SNPs form the sample. Of those five, two (LOH and VAQ) removed primarily mutations that found in both technical replicates, and three preferentially removed putative mutaions found in only one of two replicates. Two of the three filters that removed mostly non-reproducible data removed putative mutations that were within 10 bp of other putative mutations; essentially, they recognized a feature (clustered mutations) that suggested a local problem with the reads or alignment, and removed this data from consideration. A third found dbGAP overlap; essentially, found evidence that while a polymorphism may not be somatic. Our analysis suggests that these filters are useful, and do clean up the purely computaitonal data in a meaningful way.
%- LOH
By contrast, in may be more useful to discard the two filters that removed principally data from the overlap of the two replicates. Of these, one, VAQ is built into SomaticSniper. The second removes loss of heterozygosity (LOH) mutations. Multiple sources suggest that LOH mutations may be essential to cancer \cite{LOH}. This, along with the data presented here, makes a strong arugment for retaining these mutaitons in functional analyses, rather than excluding them.

%C) Address potential shortcomings. What could have gone wrong, which parts of the research might be misleading, are there any other caveats?
Although the data is suggestive, it is far from conclusive. In this analysis, we use repeatability between technical overlaps (being in the WGS and WGA samples) as a measure of confidence in a putative SNP. This metric is potentially problematic for two reasons:
%-- using repatability is also problematic 1) cancer is heterogenous; just becuase a mutaiton doesn't show up in the replicate doesn't mean it's not real
(i) First, cancer is highly heterogeneous \cite{hetero}, and so a legitimate somatic SNP might show up in one replicate and not another.
%-- using repatability is also problematic 2) could be degraded DNA, so large number of mutations could still be false positive
(ii) Second, if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates. Although repeatability does not and could not perfectly measure confidence, it should correlate. Having an independent measure of the confidence in SNP calls (repeatability), even an imperfect one, can help us gague the accuracy of other measures (filtering).  
%-- we may have shown that filtering doesn't do very much good, but we haven't really given alternatives
%-- LOH and VAQ each remove a binch of probably-real signal, but they remove a bunch of false signal too
In this sample, we have shown that while some filters function 
%-- other snp callers

@@

\begin{figure}
\includegraphics[scale=1.0]{C282_v_C484.png}
\caption{\textbf{Number of putative SNPs in WGS v. WGA, as called by SomaticSniper.} Each point is a patient. The line is y=x, so points falling below the line agree with the hypothesis that whole genome amplification makes more mutations in a sample.}
\end{figure}

\begin{figure}
\includegraphics[scale=1.0]{unfiltered_overlap_WGS_WGA_together_densities.png}
\caption{\textbf{Density of percentage overlap between WGS and WGA samples.} This figure shows the density of the percentage of each WGS (blue) and WGA (red) sample that overlaps with the other sample from the same patient. The WGS distribution is higher and narrower, showing that the WGS samples overall have a higher percentage overlap than the WGA samples, and less range in this parameter. }
\end{figure}

\begin{figure}
\includegraphics[scale=1.0]{unfiltered_total_muts_v_percent_overlap.png}
\caption{\textbf{Percent of WGS/WGA overlap versus (log) number of putative SNPs per sample.} Plot of the percentage of the WGA samples that overlaped with the corresponding WGS samples (as a measure of sample quality) against the total number of putative SNPs in the WGS sample}
\end{figure}

\begin{figure}
\includegraphics[scale=1.0]{boxplot_number_filtered.png}
\caption{\textbf{Number of putative SNPs removed by each of six filters.} The x-axis gives the name of those filters that removed putative SNPs, and the y-axis gives, on a log scale, the number of mutations removed by a given filter in a given sample. Filters removing LOH and low VAQ putative SNPs removed around 100-200 putative SNPs on average. Three other filters, removing overlap with dbSNP and putative SNPs close to other putaive mutations, removed on average ~50 mutations per sample.}
\end{figure}

\begin{figure}
% 379 samples, 311 pateints
\includegraphics[scale=1.0]{boxplot_percent_overlap_filtered.png}
\caption{\textbf{Percentage of WGS/WGA overlap removed, by filter.} The x-axis describes the filters, and the y-axis gives the percentage of the WGS/WGA overlap removed by the filter, per sample. Filters removing LOH and putative SNPs with low VAQ removed a significant ercentage of the overlap; nothing else did.}
\end{figure}

% this is not the final path name, because this is not the final figure
\begin{figure}
\includegraphics[scale=1.0]{/Users/dakota/Desktop/other_computer/LOH_all.png}
\caption{\textbf{Length of overlap of LOH segments.} As the length of the overlap between WGS and WGA increases (x-axis), the number of mutaitons filtered out by LOH increases. This is the opposite of the pattern observed in VAQ mutaitons (figure 6).}
\end{figure}

% this is not the final path name, because this is not the final figure
\begin{figure}
\includegraphics[scale=1.0]{/Users/dakota/Desktop/other_computer/VAQ_all.png}
\caption{ \textbf{Length of overlap of VAQ segments.} As the length of the overlap between WGS and WGA increases (x-axis), the number of mutaitons filtered out by VAQ decreases. This is the opposite of the pattern observed in LOH mutaitons (figure 5).}
\end{figure}
